---
title: "COVID 19 Analysis"
author: "Nicholas Cianci"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# COVID 19 Final Project

This is an exploratory data analysis project that dives into the John Hopkins
COVID-19 data set that includes Global and US case and death data 
which is publicly available on GitHub. 

To begin, I want to address some potential biases. COVID-19 data was a very
intense time for the world, myself included. Each person has their own opinion
how it was handled, what was effective vs. what was not, and what we could 
have done differently. It would be relatively easy to manipulate this data to 
support a pre-existing notion or agenda. It is important to understand and 
identify this potential upfront, so that I can proceed with every intention of 
remaining unbias in my exploration of this COVID-19 data set.

## Load Packages and Data

To begin, I installed and loaded two packages - tidyverse and lubridate to help
with my analysis throughout the project.
I then used the github links to read in the several csv files containing our
COVID-19 data.

```{r load}
#Load the tidyverse library package for use
library("tidyverse")
library("lubridate")

#Bring in the data from the github repository
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv", "time_series_covid19_confirmed_US.csv",  "time_series_covid19_deaths_US.csv")

urls<- str_c(url_in,file_names)

#Load each file into it's own object
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])

```

## Tidy Data Global Data

The Global data was split up between case data, death data, and Population data.
I decided to standardize and join them to give us one all-encompassing global
data set. I accomplished this through the use of pivots and joins.

```{r Tidy Global Data}
# Tidy Global Cases data
global_cases <- global_cases %>% 
  pivot_longer(cols = -c('Province/State',
                        'Country/Region', Lat, Long),
               names_to = "date", 
               values_to = "cases") %>%
  select(-c(Lat, Long))

# Tidy Global Death data
global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))

# Join Global cases and death data into one global data set
global <- global_cases %>% 
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

# Review summary of joined data set
summary(global)

#Standardize global data set with US data set
global <- global %>%
  unite("Combined_Key", c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE, 
        remove = FALSE)

#Read in additional file that has global population data
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat,Long_,Combined_Key,code3,iso2,iso3, Admin2))


#Join the global data set with new data that contains the global populations
global<- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths,
         Population, Combined_Key)

#View sample of newly joined global data set
tail(global)

```
## Analyze and Visualize Global Data

In this step, I analyzed and visualized the Global data set in a variety of
ways. To begin, I grouped the data so that there was 1 row for each
Country/Region that included the total number of cases, deaths, and
Country/Region Population along with two new fields which calculate the number
of cases and deaths per 1 million people. These calculated fields will help us
to see who had the highest rates of cases and deaths. If I only looked at the 
total number of cases and deaths, I would most likely only see the
Countries/Regions with the largest populations.

I used this new view of the data to filter for the Countries/Regions that had
the Top 5 Cases, Deaths, Cases per Million, and Deaths per Million. I then 
visualized each of these in their own bar chart. The visualizations were
created using ggplot and geom_bar.

```{r Global Analysis}

Global_by_country <- global %>%
  group_by(Country_Region) %>%
  summarize(cases=sum(cases), deaths = sum(deaths),
            Population=max(Population)) %>%
  mutate(cases_per_mill = cases*1000000 / Population,
         deaths_per_mill = deaths*1000000 / Population) %>%
  select(Country_Region, cases, cases_per_mill, deaths,
         deaths_per_mill, Population)

#Top 5 Cases by Country
top_5_global_cases <- Global_by_country %>%
  top_n(5,cases) %>%
  arrange(desc(cases)) %>%
  select(Country_Region,cases,Population)

top_5_global_cases %>%
  mutate(Country_Region = fct_reorder(Country_Region, cases))%>%
  ggplot(aes(y=cases,x=Country_Region, fill=Country_Region))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c("black", "blue", "green", "orange", "red"))+
  theme(legend.position="none")+
  ggtitle("Top 5 Most Global COVID-19 Cases per Country")

#Top 5 Deaths by Country
top_5_global_deaths <- Global_by_country %>%
  top_n(5,deaths) %>%
  arrange(desc(deaths)) %>%
  select(Country_Region,deaths,Population)

top_5_global_deaths %>%
  mutate(Country_Region = fct_reorder(Country_Region, deaths))%>%
  ggplot(aes(y=deaths,x=Country_Region, fill=Country_Region))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c("red4", "springgreen4", "orange", "green", "red"))+
  theme(legend.position="none")+
  ggtitle("Top 5 Most Global COVID-19 Deaths per Country")

#Top 5 Cases per Million by Country
top_5_global_cpm <- Global_by_country %>%
  top_n(5,cases_per_mill) %>%
  arrange(desc(cases_per_mill)) %>%
  select(Country_Region,cases_per_mill, Population)

top_5_global_cpm %>%
  mutate(Country_Region = fct_reorder(Country_Region, cases_per_mill))%>%
  ggplot(aes(y=cases_per_mill,x=Country_Region, fill=Country_Region))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c("blue2", "red2", "lightskyblue", "gold1", "navyblue"))+
  theme(legend.position="none")+
  ggtitle("Top 5 Highest Global COVID-19 Cases per Million per Country")

#Top 5 Deaths per Million by Country
top_5_global_dpm <- Global_by_country %>%
  top_n(5,deaths_per_mill) %>%
  arrange(desc(deaths_per_mill)) %>%
  select(Country_Region,deaths_per_mill, Population)

top_5_global_dpm %>%
  mutate(Country_Region = fct_reorder(Country_Region, deaths_per_mill))%>%
  ggplot(aes(y=deaths_per_mill,x=Country_Region, fill=Country_Region))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c("firebrick4", "darkolivegreen4", "slateblue4", "mediumseagreen", "orangered4"))+
  theme(legend.position="none")+
  ggtitle("Top 5 Most Global COVID-19 Deaths per Million per Country")
```
The Top 5 total cases and deaths bar charts were filled with large
Countries/Regions with big populations such as India, Brazil, France, Germany 
and the United States which had the highest number of both cases and deaths.

I was more interested to see which Countries/Regions had the highest cases and 
deaths per million, and it did not disappoint. These bar charts showed a much 
different picture as none of the countries with the highest totals were also in
the highest cases/deaths per million charts. 

The country with the highest cases per million was actually Australia and the
country with the highest deaths per million was Peru. It was interesting to see
smaller countries like Andorra, San Marino, and Bosnia and Herzegovina make 
appearances in those lists as well.

## Tidy US Data

I decided to do a similar process with the US data as I did with the global set.
I combined the US case and death data into one full US data set and cleared
out missing data along the way. This again was done with pivots and joins.

```{r Tidy US Data}
# Tidy US Cases data
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date=mdy(date)) %>%
  select(-c(Lat,Long_))

# Tidy US Death data
US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

# Join US cases and death data into one US data set and clear out missing data
US_total <- US_cases %>% 
  full_join(US_deaths) %>%
  filter(Province_State != "Diamond Princess")%>%
  filter(Province_State != "Grand Princess")

# Review summary of joined data set
summary(US_total)
```
## Analyze and Visualize US Data

The US data analysis followed the path of the Global data analysis with the key
difference being I grouped the data by US State rather than by country/region
and I added fields for cases/deaths per thousand rather than per million people.
I did this to account for smaller state populations.


```{r US Analysis}
# Group US data by State and add calculated fields
US_by_state <- US_total %>%
  group_by(Province_State, Country_Region,date) %>%
  summarize(cases=sum(cases),deaths=sum(deaths),
            Population =sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date, 
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

US_by_state <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(cases=sum(cases), deaths = sum(deaths),
            Population=max(Population)) %>%
  mutate(cases_per_thou = cases*1000 / Population,
         deaths_per_thou = deaths*1000 / Population) %>%
  select(Province_State, cases, cases_per_thou, deaths, deaths_per_thou, Population)

tail(US_by_state)

```

I decided to mimic the analysis and visualization performed on the global data
on the us data. Meaning, I filtered the US by State data by top 5 states with
the most cases, deaths, cases per thousand and deaths per thousand.

Again, each of these filtered views were visualized in the form of a bar chart
which was created using ggplot and geom_bar.


```{r US Analysis + Viz}

#Top 5 Cases by US State
top_5_us_cases <- US_by_state %>%
  top_n(5,cases) %>%
  select(Province_State,cases,Population)

top_5_us_cases %>%
  mutate(Province_State = fct_reorder(Province_State, cases))%>%
  ggplot(aes(y=cases,x=Province_State, fill=Province_State))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c("orange", "navyblue", "hotpink", "orange4", "gold"))+
  theme(legend.position="none")+
  ggtitle("Top 5 Most COVID-19 Cases per US State")



#Top 5 Deaths by US State
top_5_us_deaths <- US_by_state %>%
  top_n(5,deaths) %>%
  select(Province_State,deaths,Population)

top_5_us_deaths %>%
  mutate(Province_State = fct_reorder(Province_State, deaths))%>%
  ggplot(aes(y=deaths,x=Province_State, fill=Province_State))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c("chartreuse4", "hotpink", "navyblue","orange4", "gold"))+
  theme(legend.position="none")+
  ggtitle("Top 5 Most COVID-19 Deaths per US State")

#Top 5 Cases per Thousand by US State
top_5_us_cpt <- US_by_state %>%
  top_n(5,cases_per_thou) %>%
  select(Province_State,cases_per_thou, Population)

top_5_us_cpt %>%
  mutate(Province_State = fct_reorder(Province_State, cases_per_thou))%>%
  ggplot(aes(y=cases_per_thou,x=Province_State, fill=Province_State))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c("red2", "orange2", "lightskyblue3", "lightgoldenrod4", "goldenrod3"))+
  theme(legend.position="none")+
  ggtitle("Top 5 Highest COVID-19 Cases per Thousand by US State")

#Top 5 Deaths per Thousand by US State
top_5_us_dpt <- US_by_state %>%
  top_n(5,deaths_per_thou) %>%
  select(Province_State,deaths_per_thou, Population)

top_5_us_dpt %>%
  mutate(Province_State = fct_reorder(Province_State, deaths_per_thou))%>%
  ggplot(aes(y=deaths_per_thou,x=Province_State, fill=Province_State))+
  geom_bar(stat = "identity")+
  scale_fill_manual(values = c("gold4", "sienna3", "navyblue", "orangered3", "tan"))+
  theme(legend.position="none")+
  ggtitle("Top 5 Highest COVID-19 Deaths per Thousand by US State")

```

Similarly to the Global data, the states with the highest number of cases and 
deaths were states with large populations like California and Texas. 

The state with the highest deaths per thousand people was New Jersey and the
state with the highest cases per thousand people was Rhode Island. This too 
follows the pattern of the global data which saw smaller populations have higher
cases/deaths per thousand or million people. There were a few cases that didn't
quite follow this pattern such as New York which was in the highest deaths per
thousand list and Tennessee which was in the highest cases per thousand list.



## Model Creation and Visualization

I then created two models using linear regression. The first model was more to
confirm the obvious, which was that the number of cases per thousand is a good
predictor for the number of deaths per thousand. This was proven to be true 
by the visualization which shows the predicted values following the same 
pattern as the actual values on the plot. Additional evidence was provided by
the very low p-value of the model which proves that cases per thousand is a
statistically significant predictor of deaths per thousand.

The second model, I created as more of an exploratory exercise. I wanted to see
if the overall Population was a predictor of cases per thousand. I approached
this in the same way, by creating a linear regression, adding a new field
filled with predicted values and plotting the predicted values against the 
actual values. As it turns out, Population is NOT a good predictor of cases per
thousand people! We can see this in the visualization, where the predicted
values does not follow the pattern of the actual values and is further proven
by a high p-value indicating this model is not statistically significant.


```{r Models}
# Using a linear regression model to predict deaths per thousand using
# cases per thousand. 
us_mod1 <- lm(deaths_per_thou ~ cases_per_thou, data = US_by_state)
summary(us_mod1)

#Visualize actuals vs. predicted values using our first model
us_state_w_pred1 <- US_by_state %>% mutate(pred = predict(us_mod1))
us_state_w_pred1 %>% ggplot() +
  geom_point(aes(x=cases_per_thou, y=deaths_per_thou), color="blue")+
  geom_point(aes(x=cases_per_thou, y=pred), color="red")


# Using a linear regression model to predict cases per thousand using
# total state Population.
us_mod2 <- lm(cases_per_thou ~ Population, data=US_by_state)
summary(us_mod2)

#Visualize actuals vs. predicted values using our second model
us_state_w_pred2 <- US_by_state %>% mutate(pred_cpt = predict(us_mod2))
us_state_w_pred2 %>% ggplot() +
  geom_point(aes(x=Population, y=cases_per_thou), color="blue")+
  geom_point(aes(x=Population, y=pred_cpt), color="red")

```
## Conclusion

In conclusion, this was a great exercise in working with multiple data sets. 
It was certainly a challenge to load, join, tidy, analyze, visualize, and model
soooo much data. There seemed to be many different paths I could have taken. 
Ultimately, I had to choose one path as there was simply too much to work with
to analyze everything.

Throughout the analysis, it was interesting to see some initial thoughts
confirmed such as the largest countries and states having the most cases and 
deaths. It was also interesting to see some surprises come into play when 
investigating the cases/deaths per million/thousand. This was an important
lesson to learn that raw totals do not always tell the full story, often times
normalizing the data can provide a new perspective. 

Lastly, I was able to confirm one obvious suspicious using a linear regression 
model. That being that the number of cases per thousand directly predicts the
number of deaths per thousand. This makes sense and is supported visually on
our plot of actuals vs. predictions and with the low p-value of the first model.

We also learned that Population size is NOT a good predictor of cases per
thousand, at least not by using a linear regression model. While it was a
little disappointing that my guess was incorrect, it was an important part of
the project nonetheless. It goes to show that not every chart or model we make
is going to be significant. And sometimes finding what doesn't work, can help
lead us to what does.


```{r Session}

sessionInfo()

```


